---
title: "Population structure"
#output:
output:
  html_document:
    theme: flatly
    #prettydoc::html_pretty:
    #theme: cayman
    highlight: tango
    toc: true
    toc_float: true
    css: ../style.css
---


<style>
.text-box {
  background-color: #d4e9fc;
  color: black;
  font-size: 14px;
  border-radius: 5px; 
  padding: 20px
}
</style>

<style>
.lecture-box {
  background-color: #f7e1fc;
  color: black;
  font-size: 14px;
  border-radius: 5px; 
  padding: 20px
}
</style>

#<br>
#
#<div class = "lecture-box">
#
#</div>
#
#<br>

#<br>
#
#<div class = "text-box">
#
#</div>
#
#<br>


# Assessing population structure

<br>

PCA, Admixture, Fst, isolation by distance.

<br>

<div class = "lecture-box">

Short lecture on methods to detect population structure.

</div>

<br>


```{bash eval=FALSE, echo=FALSE}

```

# PCA


```bash
/plink --vcf all.vcf --out variants_pruned \
--indep-pairwise 50 5 0.2 --allow-extra-chr --double-id
```

REWRITE THIS: This command is removing SNPs in linkage using the -–indep-pairwise option. The numbers
after the option control how plink removes these SNPs. The first number is the window size (in the
number of variants it investigates). The second number is the frame-shift for the window size (how
many variants the window will move between analysis). The third number is the r2 cut-off value
(designating an upper limit for how correlated/in-linkage SNPs in a window are allowed to be). 50
5 and 0.2 are commonly used values

make the pruned file:

```bash
/plink --vcf all.vcf \
--extract variants_pruned.prune.in \
--make-bed --out variants_NoLD \
--allow-extra-chr --double-id
```

run the pca
```bash
/plink --bfile variants_NoLD \
--pca --out variants_NoLD_PCA --allow-extra-chr --double-id
```


plot the pca results
```{r, eval=F}

library(ggplot2)

setwd("~/Documents/GEOMAR/Teaching/Mar_pop_gen/2023/teaching")

dat <- read.table("variants_NoLD_PCA.eigenvec", header=F)

colnames(dat) <- c("ID", "ID2", "PC1", "PC2", "PC3", "PC4", colnames(dat)[7:ncol(dat)])

dat$population <- substr(dat$ID, 1,2)

d <- ggplot(dat, aes(PC1, PC2, fill=population)) +
        geom_point(size=4.5, shape=21, color="black") +
        #xlab(paste0("PC1: ",percentVar[1],"% variance")) +
        #ylab(paste0("PC2: ",percentVar[2],"% variance")) +
        theme_bw() +
        #scale_fill_manual(values=c('steelblue1','steelblue','grey45', "darkorchid2", "firebrick3"),
  scale_fill_manual(values=c("#D3DDDC",'#6699CC',"#F2AD00","#00A08A", "#CC3333"),
                    labels = c("Founding population","Ambient", "Acidic", 
                               "Warming", "Greenhouse"))+
        #theme(legend.title = element_blank())+
       # theme(legend.text=element_text(size=8))+
        #theme(legend.background = element_rect(colour = 'black', fill = 'white', linetype='solid'))+
#        ggtitle("F1")+
        #fill=c('steelblue1','steelblue','grey45', "darkorchid2", "firebrick3" )),order = 2),
    fill=c("#D3DDDC",'#6699CC',"#F2AD00","#00A08A", "#CC3333")),order = 2),
    shape= FALSE)

#ggsave("~/tonsa_genomics/figures/pca_afs_noF3.pdf",d, w=5.5, h=3.7)

```




# Admixture

# Fst

# Isolation by distance


add snp ids:

chr1_29988

need to get format right. its annoying.




### admixture

short lecture on population structure, Fst, drift.


```{bash, eval=F}

for K in 1 2 3 4 5 6; \
do admixture --cv variants_NoLD.bed $K | tee log${K}.out; done

```

```{r, eval=F}

samplelist <- read_tsv("~/Documents/GEOMAR/Teaching/Mar_pop_gen/2023/teaching/indivs_subsample.txt",
                       col_names = "sample")

read_delim("~/Documents/GEOMAR/Teaching/Mar_pop_gen/2023/teaching/variants_NoLD.2.Q",
                  col_names = paste0("Q",seq(1:2)),
                  delim=" ")


# read in all date, in a loop
all_data <- tibble(sample=character(),
                   k=numeric(),
                   Q=character(),
                   value=numeric())

head(all_data)

for (k in 1:6){
  data <- read_delim(paste0("~/Documents/GEOMAR/Teaching/Mar_pop_gen/2023/teaching/variants_NoLD.",k,".Q"),
                  col_names = paste0("Q",seq(1:k)),
                  delim=" ")
  data$sample <- samplelist$sample
  data$k <- k
  #This step converts from wide to long.
  data %>% gather(Q, value, -sample,-k) -> data
  all_data <- rbind(all_data,data)
}

# add the population label
all_data$population <- substr(all_data$sample, 1, 2)
all_data$population <- factor(all_data$population, 
                              levels=c("GA", "PL", "HP", "BC", "PC", "TR"))

# our orders are off in our vcf. lets re-order these from south to north. 
orderlist <- read_tsv("~/Documents/GEOMAR/Teaching/Mar_pop_gen/2023/teaching/population_order.txt",
                       col_names = "sample")
all_data$sample<-factor(all_data$sample,levels=orderlist$sample)

all_data %>%
  filter(k == 2) %>%
  ggplot(.,aes(x=sample,y=value,fill=factor(Q))) + 
  geom_rug(aes(x=sample, y=value, color=population)) +
  geom_bar(stat="identity",position="stack") +
  xlab("Sample") + ylab("Ancestry") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  scale_fill_brewer(palette="Set1",name="K",
                    labels=c("1","2"))

# plot all k values.
p <-  ggplot(all_data,aes(x=sample,y=value,fill=factor(Q))) + 
  geom_bar(stat="identity",position="stack") +
  geom_rug(aes(x=sample, color=population), inherit.aes=F) +
  xlab("Sample") + ylab("Ancestry") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 60, hjust = 1)) +
  scale_fill_brewer(palette="Set1",name="K",
                    labels=seq(1:5)) +
  facet_wrap(~k,ncol=1)



ggsave("Admixture_plot.pdf", p, width = 7, height = 15, units="in")

```


### fst

can do big windows, then selection scan?


https://owensgl.github.io/biol525D/Topic_8-9/fst.html

```bash

vcftools --gzvcf all.vcf \
--weir-fst-pop TR.pop \
--weir-fst-pop GA.pop \
--fst-window-size 25000 --out TR_GA_25kb

```

plot the pairwise fst density plots

```{r, class.source = 'fold-hide', eval=F}

# read in all in a list

take all, merge, make ridgeplot for each.
a
a
a
a
a


```

plot correlation plots

### diversity

these run decently fast. 


```{bash, eval=F}
for pop in TR PC BC HP PL GA;
do
  vcftools --vcf all.vcf \
    --keep ${pop}.pop \
    --window-pi 50000 \
    --out ${pop}_pi_50kb;
done
```

```{r, eval=F}

# read in all in a list

#take all, merge, make ridgeplot for each. 



```




## gene environment associations or selection scans.

baypass? bayenv? lfmm? pcadapt is probably easiest. 


## gwas

gemma? 

## intersection between the two?


# things to do:

consider parsing down to fewer chromosomes

Asmixture runs too slow. drop some samples? 

R libraries:
ggplot2
ggridges

drop the funny sample PC-393-024
Probably drop PL- it is confusing.

fix the error in the vcf header that gets thrown with VCFtools.
